Conditional Power and Friends: The Why and How of (Un)planned, Unblinded Sample Size Recalculations in Confirmatory Trials

### 2020-10-13 latex/manuscript.tex
lineendings
...
Updates from Overleaf

\documentclass{article}

@@

% other
\usepackage{color}
\usepackage{natbib}
\usepackage{scalerel}
\newcommand{\kr}[1]{{\color{red}#1}}

@@

Firstly, the so-called `hybrid' approach to sample size 
derivation takes a Bayesian
view on determining the initial sample size of a clinical trial~\citep{spiegelhalter1994}.
The Bayesian component requires the specification of an informative 
prior on the parameters of interest.
This then allows reasoning about the `expected power' of a trial as a
function of its sample size and, consequently, determination of the sample size
such that the expected power exceeds a target threshold. 
This concept is a straight-forward extension of the usual practice of 
computing the power under a fixed point alternative, which can easily be recovered as a special case when considering a point prior.
The advantage lies in the fact that the \textit{a~priori} information about the magnitude of the effect size is faithfully 
reflected in the sample size derivation.
Since the actual analysis is still frequentist, type-I error rate control is not compromised.

Secondly, authors have proposed to apply the concept of adaptive design changes
to recalculate the initial sample size of a design based on data observed 
within the trial itself, see e.g.~\citet{bauer2016}.
The rationale behind this approach is to use the accruing evidence about the unknown parameters driving the sample size derivation to 
`correct' the sample size mid-trial. 
During the interim analysis, the accrued data can either be unblinded
or not.
Methods for the latter `blinded interim analysis' are particularly useful if relevant nuisance parameters such as the variance are also unknown \citep{birkett1994internal,bauer2016}.
We focus on the unblinded case which allows for a more precise 
interim assessment of the effect size than methods that retain 
the blind. 
The maximal type-I error rate constraint is then usually protected by 
applying the conditional error principle~\citep{muller2004,brannath2012}
or an equivalent formulation via $\alpha$-spending or 
combination functions~\citep{bauer2016}.
Often the sample size of the current 
trial is adjusted such that the conditional power given the data observed up to the
interim analysis again exceeds the initial threshold for unconditional 
power~\citep{proschan1995}.
Any recalculation based on conditional power arguments must
address the problem that conditional power, 
just as unconditional power, 
depends on the unknown underlying parameters.
It must thus be
estimated to inform a sample size recalculation.
Yet, precise estimation of the relevant parameters before the conclusion of a trial is hard since only a fraction of the final sample size is available.
Three approaches to estimating conditional power have been discussed in the literature.

Firstly, 
`conditional power' is the probability to reject the null hypothesis given the interim data as a function of the unknown parameters.
In a slight abuse of terminology, the same term is often used to refer to
the assumed conditional power that is obtained by plugging in the 
point alternative used for the initial sample size derivation. 
Evidently, this assumed conditional makes no use of the accrued trial
data since the effect size is kept fixed.
Secondly,
authors have proposed to use the `observed conditional power' instead, which replaces the unknown parameters with their maximum-likelihood estimates.
The latter `plug-in' approach is often criticized for 
`... [using] the interim estimate of the effect [...] twice ...' \citep[p.~330]{bauer2016} (see also~\cite{bauer2006}).
Thirdly, 
conditional power can be evaluated as Bayesian expected power conditional on the observed interim data, 
i.e., by averaging conditional power as function of the unknown 
parameters with respect to the posterior density 
after conducting the interim analysis. 
Within the hybrid Bayesian framework this is usually referred to as 
`predictive power'~\citep{spiegelhalter1994,bauer2016}.

@@

Further assuming a maximal permissible type-I error rate of $\alpha=0.025$ (which we use for the remainder of the paper), the critical value for a single-stage fixed-sample-size design is the
$1-\alpha$ quantile of the standard normal distribution, i.e., approximately $c=1.96$.
The test then rejects $\mathcal{H}_0$ if and only if $Z_n > c$ after the outcomes of $n$ subjects have been observed. 
The required size of the trial $n$, given a maximal allowable type-I error
rate, is usually determined by some form of restriction on 
the (minimal) statistical power of the test.
Approaches to defining such a power constraint under 
different assumptions were reviewed in~\citet{kunzmann2020}.

@@

No matter which rationale was used to derive the overall sample size $n$,
after observing $m$, $0<m<n$, outcomes,
an independent data monitoring committee might be interested to learn about the prospects of
eventually rejecting the null hypothesis.
Both the rejection probability conditional on the presence of an effect (conditional power) or the joint probability of rejection and presence of an effect (conditional probability of success) could be of interest.
Here, we focus on the former due to the central role of
power (and conditional power) in the planning of clinical trials.
This effectively constitutes an estimation problem where the estimand is 
the \emph{conditional} probability of rejecting the null hypothesis given
$Z_m=z_m$ and $\theta > 0$.
Here the conditioning on a positive effect is crucial to justify the notion
of `power' in contrast to the unconditional `probability of success'~\citep{kunzmann2020}.

@@

The probability of rejecting the null hypothesis at the end of the trial 
as a function of $\theta$ is referred to as the `conditional power' in the literature.
It is defined as
\begin{align}
    \CP(z_m, c, \theta) :&= \Pr_\theta[\,Z_n > c \cond Z_m = z_m\,] \\
        &= 1 - \Phi\left(
            \frac{c - \sqrt{n}\,\theta - \sqrt{\tau} \, z_m + \sqrt{\tau}\sqrt{m}\,\theta
            }{\sqrt{1 - \tau}}
        \right) \\
        &= 1 - \Phi\left(
            \frac{c/\sqrt{m} + \sqrt{\tau}\,\big(\theta - \widehat{\theta}_m\big) - \theta/\sqrt{\tau}
            }{\sqrt{(n - m)/(n\,m)}}
        \right)      \ .
\end{align}
Ignoring the implicit dependency on the time point of the interim analysis, $m$, 
conditional power is thus a function of the observed effect $\widehat{\theta}_m:=z_m/\sqrt{m}$ 
(or equivalently the observed test statistic $z_m$), the critical value for the test decision~$c$, 
and the standardized effect size~$\theta$.  
Since $\theta$ is unknown, 
so is $\CP(z_m, c, \theta)$ and it thus cannot be 
evaluated directly upon observing $Z_m=z_m$.
Yet, as a function of the unknown quantity $\theta$, 
$\CP(z_m, c, \theta)$ can be estimated from observed data.
The estimation perspective on `evaluating' conditional power is 
less commonly taken in the literature but provides a consistent framework to
compare the characteristics of different methods 
(see also~\citep[Discussion]{bauer2006}).

@@

Secondly, the observed effect $\widehat{\theta}_m$ can be used 
as a plug-in estimator for the unknown effect size
(see, e.g.~\citep{proschan1995}).
This quantity is sometimes referred to as `observed conditional power'
in the literature and is defined as
\begin{align}
    \OCP(z_m, c) := \CP\big(\, z_m, c, \widehat{\theta}_m \,\big) \ .
\end{align}

Thirdly, a Bayesian approach can be taken if one 
is willing to quantify the uncertainty about the unknown parameter $\theta$ by modeling it as the
realization of a random variable $\Theta\sim\varphi(\cdot)$ where $\varphi(\theta)$ is the
prior probability density function evaluated at the parameter value $\theta$.
Our definition of this so-called `predictive power'
\begin{align}
    \PP_\varphi(z_m,c) :&= \Pr_\varphi[\,Z_n > c \cond \Theta >0, Z_m = z_m\,] \\ 
    &= \int_{0}^{\infty}\CP\big(z_m, c, \theta\big)\,\varphi\big(\theta\cond Z_m=z_m, \Theta\geq0\big)\operatorname{d}\theta
\end{align}
differs slightly from the one proposed by~\citet{spiegelhalter1994}
in that we condition on a positive effect size $\Theta>0$. 
This is more consistent with the notion of frequentist power as discussed in~\cite{kunzmann2020}, although the difference is 
only of practical relevance when a substantial
fraction of the \textit{a~priori} probability mass is concentrated on the null hypothesis.

We now compare $\ACP, \OCP$, and $\PP$ by means of a concrete example.
To this end we assume that the available prior information can be summarized in a truncated normal prior with density
\begin{align}
    \varphi(\theta) := \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\frac{\displaystyle \phi\left(\frac{\theta - 0.4}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - 0.4}{0.2}\right) - \Phi\left(\frac{-0.5 - 0.4}{0.2}\right)} \ ,
\end{align}
(see Figure~\ref{fig:prior-pdf})
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/prior_density}
    \caption{%
        Assumed prior density function. 
        The gray area indicates the null hypothesis, $\mathcal{H}_0$, of no effect. 
    }
    \label{fig:prior-pdf}
\end{figure}
and that all positive effect sizes are clinically relevant.
A truncated normal prior is convenient since it allows the analytic
computation of the posterior distribution under a normal likelihood.
It is also the maximum entropy distribution on a compact interval given mean and standard deviation and thus `least informative'
given a lower and upper boundaries on plausible effects as well as
the location and vagueness (variance) of the prior.
Following~\citep{kunzmann2020} the required 
sample size $n$ is then determined by requiring 
a minimal expected power of $1-\beta=80\%$, 
where expected power is defined as 
\begin{align}
    \EP_\varphi(n, c) :=
    \Pr\big[\,Z_n>c\,|\,\Theta\geq0\,\big] = \int \Pr_\theta\big[\,Z_n>c\,\big]\, \varphi(\theta\,|\,\Theta\geq0) \,\operatorname{d}\theta \ .
\end{align}
Under these assumptions, the required sample size is 
$n = 79$.
No point alternative is necessary to derive this sample size.
We thus use the prior mean rounded to the first decimal point to 
evaluate $\ACP$ and define $\theta_1:=0.4$.

@@

A fundamental difference between $\ACP$ and $\PP$ on the one hand and $\OCP$ on the other hand is that
$\OCP$ is the only estimator that does not condition on~$\Theta$.
In contrast, $\ACP$ implicitly conditions on $\Theta=\theta_1>0$ and 
$\PP$~on~${\Theta > 0}$.
The practical implication of this can be seen 
by looking at the sampling distributions of the 
three estimators under different assumptions 
about the value of $\theta$ (see~Figure~\ref{fig:sampling-distributions}).
Since $\OCP$ does not condition on $\Theta>0$, the sampling distribution is much more left-skewed for small effects than for the other two estimators.
Also, the negligence of the sampling variation of $\widehat{\theta}_m$, which $\OCP$ plugs in the expression for conditional power, leads to a larger variance for intermediate values of $\theta$ and the characteristic
U-shape \citep{bauer2006}.
This high variance of $\OCP$ directly translates to a relatively large mean absolute error (MAE) and mean squared error (MSE) when estimating conditional power, 
see Figure~\ref{fig:acp-ocp-pp-as-estimators}.
$\PP$ is the posterior expectation of conditional power with respect to the chosen prior and thus  minimizes the quadratic Bayes risk, i.e.~the MSE.
This leads to relative good precision for parameter values with high \textit{a~priori} likelihood 
(around $\theta=0.4$).
If one wanted to minimize the MAE directly, 
the posterior median would be optimal.
The principle, however, remains the same and
the posterior mean is more consistent with the derivation of the initial sample size using expected power. 
$\ACP$ is clearly the best in terms of bias and MAE/MSE for values close to or
above $\theta_1$, but its performance as estimator of the conditional power 
quickly deteriorates for small effect sizes.

@@

Monitoring the predictive power of an ongoing trial naturally raises the question of whether this information can be leveraged to improve the operating characteristics of a trial.
Typically, sample size recalculation is considered in the context of group-sequential designs at the pre-specified interim analysis \citep{proschan1995,brannath2004,bauer2016}.
From a statistical perspective, 
there is no reason why a sample size recalculation should not be
performed for a study that initially was planned using a single-stage design.
For the sake of simplicity the following considerations thus assume a simple 
single-stage design with fixed $n$ and $c$ as starting point and are
restricted to a single interim analysis.
All results can be generalized to the case of more complex starting designs and multiple interim analyses.

A method for sample size recalculation that is often discussed in the literature is to derive a new sample size $n'$ and a new critical value $c'$ such that some estimate of conditional power exceeds a threshold $1-\beta_{cond.}$.
Different choices for the estimator of conditional power
and the choice of $\beta_{cond.}$ have been discussed in the literature \citep{proschan1995,brannath2004,bauer2016}.
Strict overall type-I error rate control can be maintained by invoking the conditional error principle \citep{muller2001,muller2004,brannath2012}, i.e., by limiting the maximal type-I error rate of the new design to the maximal conditional type-I error rate of the original design given the observed $z_m$.
Often, trial protocols leave the exact choice of $1-\beta_{cond.}$ open since it can be chosen \textit{ad~hoc}
without compromising strict type-I-error rate control.
The basic concept of readjusting the sample size to achieve the desired conditional power is, however, a common approach, see for instance,~\citet{bhatt2013} and~\citet{mehta2009}.

@@

Similar to suggestions in the literature on binding futility stopping, 
one could modify the initial nominal $\alpha$ and the conditional power threshold $\beta_{cond.}$ until the recalculated design again fulfills the required 
operating characteristics \citep{brannath2004}. 
This approach to define a fully pre-specified two-stage design is, however, unnecessarily complicated in that it 
uses methods like the conditional error principle or a combination function, which are originally intended for \emph{unplanned}
design adaptations.

@@

Instead, the interim time-point $m$, 
the sample size function $n(z_m)$, 
the critical value function $c(z_m)$,
and the early futility- and efficacy boundaries 
($n(z_m)=m$ and $ c(z_m)=\pm\infty$)
can be optimized directly.
For the sample size function $n(z_m)$ \citet{brannath2004} discussed this approach by fitting a polynomial of degree 4 but they did not optimize
any of the other parameters.
In a more general form, \citet{pilz2019} approached the problem using
variational methods.
An \texttt{R} implementation that uses cubic splines for both the $n$ and
$c$ functions is available via the package~\textit{adoptr}~\citep{adoptrjss2020}.

In the following we restrict the considerations to the case
where only expected sample size is of interest and the variation is
ignored, i.e.~$\eta=0$.
The corresponding optimal two-stage design is the solution of
\begin{align}
    \argmin{m, n(\cdot), c(\cdot)} 
        &&                                   \E_\varphi[\,n(Z_m)\,] & \label{eq:optimal-design-start}\\
    \st &&                 \Pr_{0}[\,Z_{n(Z_m)} > c(Z_m)\,] & \leq \alpha \\
        &&     \EP_\varphi\big(n(\cdot), c(\cdot)\big) & \geq 1 - \beta \label{eq:optimal-design-end}
\end{align}
and can be derived numerically using the \texttt{R} package \texttt{adoptr}~\citep{adoptrjss2020}.
Here, $\E_\varphi[\,n(Z_m)\,]$ is the expected sample size under the prior $\varphi$ and expected power needs to be redefined to account for the
fact that $n$ and $c$ are now functions of the interim results
\begin{align}
    \EP_\varphi\big(n(\cdot), c(\cdot)\big) :=
    \Pr_\varphi\big[\,Z_{n(Z_m)}>c(Z_m)\,|\,\Theta\geq0\,\big] \ .
\end{align}
Figure~\ref{fig:binding-recalculation} shows the sample size-, critical value-, and the predictive power function in the example situation discussed earlier together with the single-stage design
and the na\"{\i}ve adaptation based on predictive power discussed in
Section~\ref{sec:naive-adaptations}.
The optimal two-stage design complies with both the predictive power and the maximal type-I error rate constraints and is thus comparable with the original single-stage design in terms of sample size.
The expected sample size is much lower ($56.4$ \emph{versus} $79$) at the cost of a non-zero standard deviation of the expected sample size ($28.5$ \emph{versus} $0$).
Since it is entirely pre-specified, 
the optimal two-stage design does not need to fall back to the conditional error principle to control the maximal type-I error rate.
Instead, it achieves the desired design characteristics in an optimal way since they are incorporated as constraints to problem \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
The optimal design fully exhausts the allowable maximal type-I error rate
and complies with the expected power constraint. 
This comes at the cost of a predictive power that can drop as low as 40\% close to the futility boundary
as shown in the right panel of Figure~\ref{fig:binding-recalculation}.
The sample size function of the optimal design is also 
characteristically different from the na\"{\i}ve design.
The latter is convex on the continuation region whereas the optimal
shape has a mode close to the early-futility boundary.
A recalculation based on exceeding a fixed lower threshold for 
some estimator of conditional power always results in a convex
sample size function and can thus never be fully optimal irrespective of how the nominal values of $\alpha$ and $\beta_{cond.}$
are chosen.

@@

A recalculated sample size under a different prior $\psi$ can then be obtained by 
keeping $\lambda_h$ fixed and minimizing 
\begin{align}
    \argmin{n',c'} && n'&-\lambda_h\, h_\psi(n',c') \\
     \st && g(n', c') &\leq 0 \ . 
\end{align}
This effectively relaxes the predictive power constraint by allowing the same trade-off 
between sample size and predictive power that was implied by the original optimal design.
The conditional type-I error rate constraint is not relaxed to comply with the 
conditional error principle.
This approach is closely related to the formulation of the optimal-design problem given in~\citet{jennison2015} and the utility perspective on sample 
size derivation presented in~\citep{kunzmann2020}.
In contrast to the situation analyzed by \citet{jennison2015}, however,
$\lambda_h$ does depend on $z_m$.
This is due to the fact that expected sample size is formed with respect to the unconditional
prior while expected power uses the prior conditional on $\Theta>0$.

@@

A limitation of the proposed recalculation methods is the fact that they 
rely on the conditional error principle for strict type-I-error rate control.
The conditional error principle can be difficult to extend to cases with nuisance
parameters \citep{gutjahr2011}.
In practice, a simple plug-in approach similar to the approach in blinded sample size reassessment might be viable but have not yet been investigated more thoroughly.
Alternatively, any other form of pre-specifying a combination function for the stage-wise p values might be used to control the type-I-error rate.
This combination function should then also be optimized over during the planning stage 
to avoid inconsistencies. 
In particular, the optimal combination function of a two stage-design minimizing expected sample size can be approximated with an inverse normal combination test~\citep{pilz2019}.

\bibliographystyle{plainnat}
\bibliography{references.bib}

>>>>

\documentclass[12pt]{article}

@@

% other
\usepackage{color}
\usepackage[numbers,super,sort&compress]{NJDnatbib}
\bibliographystyle{WileyNJD-AMA}
\usepackage{scalerel}
\newcommand{\kr}[1]{{\color{red}#1}}

@@

Firstly, the so-called `hybrid' approach to sample size
derivation takes a Bayesian
view on determining the initial sample size of a clinical trial~\cite{spiegelhalter1994}.
The Bayesian component requires the specification of an informative
prior on the parameters of interest.
This then allows reasoning about the `expected power' of a trial as a
function of its sample size and, consequently, determination of the sample size
such that the expected power exceeds a target threshold.
This concept is a straight-forward extension of the usual practice of
computing the power under a fixed point alternative, which can easily be recovered as a special case when considering a point prior.
The advantage lies in the fact that the \textit{a~priori} information about the magnitude of the effect size is faithfully
reflected in the sample size derivation.
Since the actual analysis is still frequentist, type-I error rate control is not compromised.

Secondly, authors have proposed to apply the concept of adaptive design changes
to recalculate the initial sample size of a design based on data observed
within the trial itself~\cite{bauer2016}.
The rationale behind this approach is to use the accruing evidence about the unknown parameters driving the sample size derivation to
`correct' the sample size mid-trial.
During the interim analysis, the accrued data can either be unblinded
or not.
Methods for the latter `blinded interim analysis' are particularly useful if relevant nuisance parameters such as the variance are also unknown \cite{birkett1994internal,bauer2016}.
We focus on the unblinded case which allows for a more precise
interim assessment of the effect size than methods that retain
the blind.
The maximal type-I error rate constraint is then usually protected by
applying the conditional error principle~\cite{muller2004,brannath2012}
or an equivalent formulation via $\alpha$-spending or
combination functions~\cite{bauer2016}.
Often the sample size of the current
trial is adjusted such that the conditional power given the data observed up to the
interim analysis again exceeds the initial threshold for unconditional
power~\cite{proschan1995}.
Any recalculation based on conditional power arguments must
address the problem that conditional power,
just as unconditional power,
depends on the unknown underlying parameters.
It must thus be
estimated to inform a sample size recalculation.
Yet, precise estimation of the relevant parameters before the conclusion of a trial is hard since only a fraction of the final sample size is available.
Three approaches to estimating conditional power have been discussed in the literature.

Firstly,
`conditional power' is the probability to reject the null hypothesis given the interim data as a function of the unknown parameters.
In a slight abuse of terminology, the same term is often used to refer to
the assumed conditional power that is obtained by plugging in the
point alternative used for the initial sample size derivation.
Evidently, this assumed conditional makes no use of the accrued trial
data since the effect size is kept fixed.
Secondly,
authors have proposed to use the `observed conditional power' instead, which replaces the unknown parameters with their maximum-likelihood estimates.
The latter `plug-in' approach is often criticized for
`... [using] the interim estimate of the effect [...] twice ...' Bauer~\textit{et~al.},~p.~330~\cite{bauer2016} and \cite{bauer2006}.
Thirdly,
conditional power can be evaluated as Bayesian expected power conditional on the observed interim data,
i.e., by averaging conditional power as function of the unknown
parameters with respect to the posterior density
after conducting the interim analysis.
Within the hybrid Bayesian framework this is usually referred to as
`predictive power'~\cite{spiegelhalter1994,bauer2016}.

@@

Further assuming a maximal permissible type-I error rate of $\alpha=0.025$ (which we use for the remainder of the paper), the critical value for a single-stage fixed-sample-size design is the
$1-\alpha$ quantile of the standard normal distribution, i.e., approximately $c=1.96$.
The test then rejects $\mathcal{H}_0$ if and only if $Z_n > c$ after the outcomes of $n$ subjects have been observed.
The required size of the trial $n$, given a maximal allowable type-I error
rate, is usually determined by some form of restriction on
the (minimal) statistical power of the test.
Approaches to defining such a power constraint under
different assumptions were reviewed in~\cite{kunzmann2020}.

@@

No matter which rationale was used to derive the overall sample size $n$,
after observing $m$, $0<m<n$, outcomes,
an independent data monitoring committee might be interested to learn about the prospects of
eventually rejecting the null hypothesis.
Both the rejection probability conditional on the presence of an effect (conditional power) or the joint probability of rejection and presence of an effect (conditional probability of success) could be of interest.
Here, we focus on the former due to the central role of
power (and conditional power) in the planning of clinical trials.
This effectively constitutes an estimation problem where the estimand is
the \emph{conditional} probability of rejecting the null hypothesis given
$Z_m=z_m$ and $\theta > 0$.
Here the conditioning on a positive effect is crucial to justify the notion
of `power' in contrast to the unconditional `probability of success'~\cite{kunzmann2020}.

@@

The probability of rejecting the null hypothesis at the end of the trial
as a function of $\theta$ is referred to as the `conditional power' in the literature.
It is defined as
\begin{align}
    \CP(z_m, c, \theta) :&= \Pr_\theta[\,Z_n > c \cond Z_m = z_m\,] \\
        &= 1 - \Phi\left(
            \frac{c - \sqrt{n}\,\theta - \sqrt{\tau} \, z_m + \sqrt{\tau}\sqrt{m}\,\theta
            }{\sqrt{1 - \tau}}
        \right) \\
        &= 1 - \Phi\left(
            \frac{c/\sqrt{m} + \sqrt{\tau}\,\big(\theta - \widehat{\theta}_m\big) - \theta/\sqrt{\tau}
            }{\sqrt{(n - m)/(n\,m)}}
        \right)      \ .
\end{align}
Ignoring the implicit dependency on the time point of the interim analysis, $m$,
conditional power is thus a function of the observed effect $\widehat{\theta}_m:=z_m/\sqrt{m}$
(or equivalently the observed test statistic $z_m$), the critical value for the test decision~$c$,
and the standardized effect size~$\theta$.
Since $\theta$ is unknown,
so is $\CP(z_m, c, \theta)$ and it thus cannot be
evaluated directly upon observing $Z_m=z_m$.
Yet, as a function of the unknown quantity $\theta$,
$\CP(z_m, c, \theta)$ can be estimated from observed data.
The estimation perspective on `evaluating' conditional power is
less commonly taken in the literature but provides a consistent framework to
compare the characteristics of different methods~\cite{bauer2006}.

@@

Secondly, the observed effect $\widehat{\theta}_m$ can be used
as a plug-in estimator for the unknown effect size~\cite{proschan1995}.
This quantity is sometimes referred to as `observed conditional power'
in the literature and is defined as
\begin{align}
    \OCP(z_m, c) := \CP\big(\, z_m, c, \widehat{\theta}_m \,\big) \ .
\end{align}

Thirdly, a Bayesian approach can be taken if one
is willing to quantify the uncertainty about the unknown parameter $\theta$ by modeling it as the
realization of a random variable $\Theta\sim\varphi(\cdot)$ where $\varphi(\theta)$ is the
prior probability density function evaluated at the parameter value $\theta$.
Our definition of this so-called `predictive power'
\begin{align}
    \PP_\varphi(z_m,c) :&= \Pr_\varphi[\,Z_n > c \cond \Theta >0, Z_m = z_m\,] \\
    &= \int_{0}^{\infty}\CP\big(z_m, c, \theta\big)\,\varphi\big(\theta\cond Z_m=z_m, \Theta\geq0\big)\operatorname{d}\theta
\end{align}
differs slightly from the one proposed by~\cite{spiegelhalter1994}
in that we condition on a positive effect size $\Theta>0$.
This is more consistent with the notion of frequentist power as discussed in~\cite{kunzmann2020}, although the difference is
only of practical relevance when a substantial
fraction of the \textit{a~priori} probability mass is concentrated on the null hypothesis.

We now compare $\ACP, \OCP$, and $\PP$ by means of a concrete example.
To this end we assume that the available prior information can be summarized in a truncated normal prior with density
\begin{align}
    \varphi(\theta) := \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\frac{\displaystyle \phi\left(\frac{\theta - 0.4}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - 0.4}{0.2}\right) - \Phi\left(\frac{-0.5 - 0.4}{0.2}\right)} \ ,
\end{align}
(see Figure~\ref{fig:prior-pdf})
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/prior_density}
    \caption{%
        Assumed prior density function.
        The gray area indicates the null hypothesis, $\mathcal{H}_0$, of no effect.
    }
    \label{fig:prior-pdf}
\end{figure}
and that all positive effect sizes are clinically relevant.
A truncated normal prior is convenient since it allows the analytic
computation of the posterior distribution under a normal likelihood.
It is also the maximum entropy distribution on a compact interval given mean and standard deviation and thus `least informative'
given a lower and upper boundaries on plausible effects as well as
the location and vagueness (variance) of the prior.
Following~\cite{kunzmann2020} the required
sample size $n$ is then determined by requiring
a minimal expected power of $1-\beta=80\%$,
where expected power is defined as
\begin{align}
    \EP_\varphi(n, c) :=
    \Pr\big[\,Z_n>c\,|\,\Theta\geq0\,\big] = \int \Pr_\theta\big[\,Z_n>c\,\big]\, \varphi(\theta\,|\,\Theta\geq0) \,\operatorname{d}\theta \ .
\end{align}
Under these assumptions, the required sample size is
$n = 79$.
No point alternative is necessary to derive this sample size.
We thus use the prior mean rounded to the first decimal point to
evaluate $\ACP$ and define $\theta_1:=0.4$.

@@

A fundamental difference between $\ACP$ and $\PP$ on the one hand and $\OCP$ on the other hand is that
$\OCP$ is the only estimator that does not condition on~$\Theta$.
In contrast, $\ACP$ implicitly conditions on $\Theta=\theta_1>0$ and
$\PP$~on~${\Theta > 0}$.
The practical implication of this can be seen
by looking at the sampling distributions of the
three estimators under different assumptions
about the value of $\theta$ (see~Figure~\ref{fig:sampling-distributions}).
Since $\OCP$ does not condition on $\Theta>0$, the sampling distribution is much more left-skewed for small effects than for the other two estimators.
Also, the negligence of the sampling variation of $\widehat{\theta}_m$, which $\OCP$ plugs in the expression for conditional power, leads to a larger variance for intermediate values of $\theta$ and the characteristic
U-shape \cite{bauer2006}.
This high variance of $\OCP$ directly translates to a relatively large mean absolute error (MAE) and mean squared error (MSE) when estimating conditional power,
see Figure~\ref{fig:acp-ocp-pp-as-estimators}.
$\PP$ is the posterior expectation of conditional power with respect to the chosen prior and thus  minimizes the quadratic Bayes risk, i.e.~the MSE.
This leads to relative good precision for parameter values with high \textit{a~priori} likelihood
(around $\theta=0.4$).
If one wanted to minimize the MAE directly,
the posterior median would be optimal.
The principle, however, remains the same and
the posterior mean is more consistent with the derivation of the initial sample size using expected power.
$\ACP$ is clearly the best in terms of bias and MAE/MSE for values close to or
above $\theta_1$, but its performance as estimator of the conditional power
quickly deteriorates for small effect sizes.

@@

Monitoring the predictive power of an ongoing trial naturally raises the question of whether this information can be leveraged to improve the operating characteristics of a trial.
Typically, sample size recalculation is considered in the context of group-sequential designs at the pre-specified interim analysis \cite{proschan1995,brannath2004,bauer2016}.
From a statistical perspective,
there is no reason why a sample size recalculation should not be
performed for a study that initially was planned using a single-stage design.
For the sake of simplicity the following considerations thus assume a simple
single-stage design with fixed $n$ and $c$ as starting point and are
restricted to a single interim analysis.
All results can be generalized to the case of more complex starting designs and multiple interim analyses.

A method for sample size recalculation that is often discussed in the literature is to derive a new sample size $n'$ and a new critical value $c'$ such that some estimate of conditional power exceeds a threshold $1-\beta_{cond.}$.
Different choices for the estimator of conditional power
and the choice of $\beta_{cond.}$ have been discussed in the literature \cite{proschan1995,brannath2004,bauer2016}.
Strict overall type-I error rate control can be maintained by invoking the conditional error principle \cite{muller2001,muller2004,brannath2012}, i.e., by limiting the maximal type-I error rate of the new design to the maximal conditional type-I error rate of the original design given the observed $z_m$.
Often, trial protocols leave the exact choice of $1-\beta_{cond.}$ open since it can be chosen \textit{ad~hoc}
without compromising strict type-I-error rate control.
The basic concept of readjusting the sample size to achieve the desired conditional power is, however, a common approach, see for instance,~\cite{bhatt2013} and~\cite{mehta2009}.

@@

Similar to suggestions in the literature on binding futility stopping,
one could modify the initial nominal $\alpha$ and the conditional power threshold $\beta_{cond.}$ until the recalculated design again fulfills the required
operating characteristics \cite{brannath2004}.
This approach to define a fully pre-specified two-stage design is, however, unnecessarily complicated in that it
uses methods like the conditional error principle or a combination function, which are originally intended for \emph{unplanned}
design adaptations.

@@

Instead, the interim time-point $m$,
the sample size function $n(z_m)$,
the critical value function $c(z_m)$,
and the early futility- and efficacy boundaries
($n(z_m)=m$ and $ c(z_m)=\pm\infty$)
can be optimized directly.
For the sample size function $n(z_m)$ \cite{brannath2004} discussed this approach by fitting a polynomial of degree 4 but they did not optimize
any of the other parameters.
In a more general form, \cite{pilz2019} approached the problem using
variational methods.
An \texttt{R} implementation that uses cubic splines for both the $n$ and
$c$ functions is available via the package~\textit{adoptr}~\cite{adoptrjss2020}.

In the following we restrict the considerations to the case
where only expected sample size is of interest and the variation is
ignored, i.e.~$\eta=0$.
The corresponding optimal two-stage design is the solution of
\begin{align}
    \argmin{m, n(\cdot), c(\cdot)}
        &&                                   \E_\varphi[\,n(Z_m)\,] & \label{eq:optimal-design-start}\\
    \st &&                 \Pr_{0}[\,Z_{n(Z_m)} > c(Z_m)\,] & \leq \alpha \\
        &&     \EP_\varphi\big(n(\cdot), c(\cdot)\big) & \geq 1 - \beta \label{eq:optimal-design-end}
\end{align}
and can be derived numerically using the \texttt{R} package \texttt{adoptr}~\cite{adoptrjss2020}.
Here, $\E_\varphi[\,n(Z_m)\,]$ is the expected sample size under the prior $\varphi$ and expected power needs to be redefined to account for the
fact that $n$ and $c$ are now functions of the interim results
\begin{align}
    \EP_\varphi\big(n(\cdot), c(\cdot)\big) :=
    \Pr_\varphi\big[\,Z_{n(Z_m)}>c(Z_m)\,|\,\Theta\geq0\,\big] \ .
\end{align}
Figure~\ref{fig:binding-recalculation} shows the sample size-, critical value-, and the predictive power function in the example situation discussed earlier together with the single-stage design
and the na\"{\i}ve adaptation based on predictive power discussed in
Section~\ref{sec:naive-adaptations}.
The optimal two-stage design complies with both the predictive power and the maximal type-I error rate constraints and is thus comparable with the original single-stage design in terms of sample size.
The expected sample size is much lower ($56.4$ \emph{versus} $79$) at the cost of a non-zero standard deviation of the expected sample size ($28.5$ \emph{versus} $0$).
Since it is entirely pre-specified,
the optimal two-stage design does not need to fall back to the conditional error principle to control the maximal type-I error rate.
Instead, it achieves the desired design characteristics in an optimal way since they are incorporated as constraints to problem \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
The optimal design fully exhausts the allowable maximal type-I error rate
and complies with the expected power constraint.
This comes at the cost of a predictive power that can drop as low as 40\% close to the futility boundary
as shown in the right panel of Figure~\ref{fig:binding-recalculation}.
The sample size function of the optimal design is also
characteristically different from the na\"{\i}ve design.
The latter is convex on the continuation region whereas the optimal
shape has a mode close to the early-futility boundary.
A recalculation based on exceeding a fixed lower threshold for
some estimator of conditional power always results in a convex
sample size function and can thus never be fully optimal irrespective of how the nominal values of $\alpha$ and $\beta_{cond.}$
are chosen.

@@

A recalculated sample size under a different prior $\psi$ can then be obtained by
keeping $\lambda_h$ fixed and minimizing
\begin{align}
    \argmin{n',c'} && n'&-\lambda_h\, h_\psi(n',c') \\
     \st && g(n', c') &\leq 0 \ .
\end{align}
This effectively relaxes the predictive power constraint by allowing the same trade-off
between sample size and predictive power that was implied by the original optimal design.
The conditional type-I error rate constraint is not relaxed to comply with the
conditional error principle.
This approach is closely related to the formulation of the optimal-design problem given in~\cite{jennison2015} and the utility perspective on sample
size derivation presented in~\cite{kunzmann2020}.
In contrast to the situation analyzed by \cite{jennison2015}, however,
$\lambda_h$ does depend on $z_m$.
This is due to the fact that expected sample size is formed with respect to the unconditional
prior while expected power uses the prior conditional on $\Theta>0$.

@@

A limitation of the proposed recalculation methods is the fact that they
rely on the conditional error principle for strict type-I-error rate control.
The conditional error principle can be difficult to extend to cases with nuisance
parameters \cite{gutjahr2011}.
In practice, a simple plug-in approach similar to the approach in blinded sample size reassessment might be viable but have not yet been investigated more thoroughly.
Alternatively, any other form of pre-specifying a combination function for the stage-wise p values might be used to control the type-I-error rate.
This combination function should then also be optimized over during the planning stage
to avoid inconsistencies.
In particular, the optimal combination function of a two stage-design minimizing expected sample size can be approximated with an inverse normal combination test~\cite{pilz2019}.

\bibliography{manuscript.bib}
